Recall that if $f$ is a map and $A$ is a set, the {\em restriction} of $f$ to $A$ is the set:
    \[
    f_{|A}=\{(x,y): x\in A\mbox{\ and\ }(x,y)\in f\}
    \]
This is also a set of ordered pairs such that $y=y'$ whenever
$(x,y)\in f_{|A}$ and $(x,y')\in f_{|A}$. Hence the restriction
$f_{|A}$ is also a map. Furthermore if $f:A'\to B$ and $A\subseteq
A'$, then $f_{|A}:A\to B$.
\index{free@Free generator of universal algebra}
\begin{defin}\label{logic:def:free:generator}
Let $X$ be a universal algebra of type $\alpha$. We call {\em free
generator} of $X$ any subset $X_{0}\subseteq X$ with the following
property: for any universal algebra $Y$ of type $\alpha$, and for
any map $g_{0}:X_{0}\to Y$, there exists a unique morphism $g:X\to
Y$ such that $g_{|X_{0}}=g_{0}$.
\end{defin}
Consider the set $X=\N$ together with the successor operator
$s:\N\to\N$ defined by $s(n)=n+1$. Then $X$ can be viewed as a
universal algebra of type $\alpha=\{(0,1)\}$ where $T(0,1):X^{1}\to
X$ is defined as $T(0,1)(n)=n(0)+1$. Furthermore the set
$X_{0}=\{0\}$ is a free generator of $X$. To check this, let $Y$ be
another universal algebra of type $\alpha$ and assume that
$g_{0}:X_{0}\to Y$ is an arbitrary map. We have to prove that
$g_{0}$ can be uniquely extended into a morphism $g:X\to Y$. This is
another way of saying that there exists a morphism $g:X\to Y$ such
that $g_{|X_{0}}=g_{0}$, and that such morphism is unique. Since
$X_{0}=\{0\}$, we have $g_{0}:\{0\}\to Y$. We define $e=g_{0}(0)\in
Y$. Consider the map $g:X\to Y$ defined by the recursion $g(0)=e$
and $g(n+1) = s\circ g(n)$, where $s:Y^{1}\to Y$ is the successor
operator on $Y$. To check that $g$ is a morphism, we need to check
the single condition for all $n\in X$:
    \[
    g\circ s(n) = s\circ g(n)
    \]
which is true from the very definition of $g$. Hence we have proved
the existence of a morphism $g:X\to Y$, and since $g(0) = e =
g_{0}(0)$ we have $g_{|X_{0}}=g_{0}$. It remains to check that $g$
is unique. Suppose that $g':X\to Y$ is another morphism such that
$g'_{|X_{0}}=g_{0}$, i.e. $g'(0)=e$. In particular $g'(0) = g(0)$.
Furthermore, since $g'$ is a morphism, if $g'(n)=g(n)$ for some
$n\in X$, then:
    \[
    g'(n+1) = s\circ g'(n) =  s\circ g(n) = g(n+1)
    \]
This proves by induction that $g'(n)=g(n)$ for all $n\in X$, i.e.
that $g'=g$.

We have proved that $\{0\}$ is a free generator of $\N$ viewed as a
universal algebra with a single successor operator. Note that the
empty set alone would be too small to be a free generator of $\N$.
If $g_{0}:\emptyset\to Y$ is a map then $g_{0}$ has to be the empty
set, and it is not difficult to extend $g_{0}$ into a full morphism
$g:X\to Y$ whenever $Y$ is not empty. Indeed pick an arbitrary $e\in
Y$ and define $g$ by the same recursion as before, $g(0)=e$ and
$g(n+1)=s\circ g(n)$. We obtain a morphism such that
$g_{|X_{0}}=\emptyset=g_{0}$. However, such morphism will not be
unique as there are in general many ways to choose the element $e\in
Y$. This shows that the empty set is not a free generator of $\N$
viewed as a universal algebra with a single successor operator. In
fact, we could have made the argument simpler: consider
$Y=\emptyset$ which is a universal algebra of type $\alpha$ by
setting $T(f)=\emptyset$ for all $f\in\alpha$\footnote{We have
$T(f):Y^{\alpha(f)}\to Y$ since $\alpha(f)\geq 1$ for all
$f\in\alpha$. The empty set $Y=\emptyset$ is a universal algebra of
type $\alpha$ only when $\alpha$ has no constant, i.e. no operator
with arity $0$.}. Consider the map $g_{0}:\emptyset\to Y$ which is
just the empty map. Then it is impossible to extend $g_{0}$ to a
full morphism $g:X\to Y$, for the simple reason that there cannot
exist any map $g:X\to\emptyset$ since $X$ is not empty.

We now claim that $X_{0}=\{0,1\}$ is too large to be a free
generator of $\N$. In the case of $X_{0}=\emptyset$, we failed to
obtain uniqueness when $Y$ had more than one element, and we failed
to obtain existence for $Y=\emptyset$. In the case of
$X_{0}=\{0,1\}$, we will fail to obtain existence. Indeed consider
$Y=\N$ and $g_{0}:X_{0}\to Y$ defined by $g_{0}(0)=2$ and
$g_{0}(1)=4$. Then no morphism $g:X\to Y$ can extend $g_{0}$, as
otherwise:
    \[
    4 = g_{0}(1)=g(1)=g(0+1)=g(0)+1 = g_{0}(0) + 1 = 3
    \]

So we have seen that $X_{0}=\emptyset$ is too small, while
$X_{0}=\{0,1\}$ is too large to be a free generator of $\N$. This
last statement is in fact misleading as size is not everything.
Consider $X_{0}=\{1\}$ with $Y=\N$ and $g_{0}:X_{0}\to Y$ defined by
$g_{0}(1)=0$. Then no morphism $g:X\to Y$ can extend $g_{0}$ as
otherwise:
    \[
    0 = g_{0}(1)=g(1)=g(0+1)=g(0)+1
    \]
contradicting the fact that $g(0)\in\N$. So $X_{0}=\{1\}$ is not a
free generator of $\N$ viewed as a universal algebra with successor
operator.

Suppose we now regard $\N$ as a universal algebra of type
$\alpha=\{(0,1),(1,0)\}$ by adding to the successor operator the
operator $T(1,0):\{0\}\to\N$ defined by $T(1,0)(0) = 0$. In effect,
we are giving $\N$ additional structure by singling out the element
$0$. We now claim that the empty set $X_{0}=\emptyset$ is a free
generator of $\N$. So let $Y$ be a universal algebra of type
$\alpha$ and consider a map $g_{0}:X_{0}\to Y$. Since
$X_{0}=\emptyset$, $g_{0}$ is necessarily the empty map
$g_{0}=\emptyset$. We need to show that $g_{0}$ can be uniquely
extended into a morphism $g:X\to Y$. Since $Y$ is a universal
algebra of type $\alpha$, we have an operator $T(1,0):\{0\}\to Y$.
Define $e=T(1,0)(0)\in Y$ and consider $g:X\to Y$ defined by the
recursion $g(0)=e$ and $g(n+1)=s\circ g(n)$. Then $g$ is a morphism
as it satisfies the two conditions:
    \[
    g\circ T(1,0)(0) = g(0) = e = T(1,0)(0) =T(1,0)\circ g^{0}(0)
    \]
and:
    \[
    g\circ T(0,1)(n) = g\circ s(n)= g(n+1)=s\circ g(n) = T(0,1)\circ g^{1}(n)
    \]
Furthermore, $g_{|X_{0}}=\emptyset = g_{0}$, and since any other
morphism $g':X\to Y$ would need to satisfy $g'(0)=e$ and
$g'(n+1)=s\circ g'(n)$, a simple induction argument shows that $g$
is in fact unique. So we have proved that $X_{0}=\emptyset$ is a
free generator of $\N$ viewed as a universal algebra of type
$\alpha$.

Before we end this section, let us confront our fear of the empty
set once more. We have just seen that it is possible for
$X_{0}=\emptyset$ to be a free generator of a universal algebra
which is not empty. But such universal algebra had a constant $0$.
Suppose now that $\alpha$ is a type of universal algebra without
constant, i.e. such that $\alpha(f)\geq 1$ for all $f\in\alpha$.
Then the empty set $X=\emptyset$ is a universal algebra of type
$\alpha$, provided the map $T$ of domain $\alpha$ is defined as
$T(f)=\emptyset$ for all $f\in\alpha$. Then $T(f)$ is indeed a map
$T(f):X^{\alpha(f)}\to X$, and it is the only possible choice. We
claim that $X$ is the only universal algebra of type $\alpha$ which
can have $X_{0}=\emptyset$ as a free generator. So first we show
that $X_{0}=\emptyset$ is indeed a free generator of $X$. Suppose
$Y$ is a universal algebra of type $\alpha$ and let $g_{0}:X_{0}\to
Y$ be a map. Then $g_{0}$ is necessarily the empty map. Define
$g=g_{0}=\emptyset$. Then $g:X\to Y$ and in fact $g$ is a morphism
since the conditions required $\forall f\in\alpha\ ,\ \forall x\in
X^{\alpha(f)}\ldots$ are vacuously satisfied, as
$X^{\alpha(f)}=\emptyset$ whenever $X=\emptyset$ and $\alpha(f)\geq
1$. We clearly have $g_{|X_{0}}=\emptyset =g_{0}$. Furthermore, the
morphism $g=\emptyset$ is unique, since it is the only possible map
$g:X\to Y$. So we have proved that $X_{0}=\emptyset$ is a free
generator of $X=\emptyset$ viewed as a universal algebra of type
$\alpha$. We now show that $X=\emptyset$ is the only universal
algebra of type $\alpha$ which can have $X_{0}=\emptyset$ as a free
generator. So suppose $X_{0}=\emptyset$ is a free generator of some
universal algebra $X$ of type $\alpha$. Let $Y=\emptyset$ be viewed
as a universal algebra of type $\alpha$ with the obvious structure
$T$. Consider the empty map $g_{0}:X_{0}\to Y$. Since $X_{0}$ is a
free generator, $g_{0}$ can be extended into a morphism $g:X\to Y$.
But since $Y$ is empty, there exists no map $g:X\to Y$, unless $X$
is also empty. So we have proved that $X=\emptyset$.
