\begin{thebibliography}{99}
\bibitem{AndrekaNemeti}
Hajnal Andr\'eka, Istv\'an N\'emeti, Ildik\'o Sain, (2013).
Universal Algebraic Logic. Studies in Universal Logic
\bibitem{Cylindric}
H. Andr\'eka, I. N\'emeti, L. Henkin, J.D. Monk, A. Tarski (1981).
Cylindric Set Algebras. Lecture Notes in Mathematics. Springer.
\bibitem{ComComp}
Sanjeev Arora, Boaz Barak, (2009). Computational Complexity.
Cambridge University Press.
\bibitem{AlgLog}
Donald W. Barnes, John M. Mack, (1975). An Algebraic Introduction to
Mathematical Logic. Graduate texts in mathematics. Springer.
\bibitem{Bell}
John L. Bell, (2005). Set Theory. Oxford Logic Guides.
\bibitem{BlokPigozzi}
Blok W.J., Pigozzi D., (1989), Algebraizable Logics, Memoirs of the
American Mathematical Society, Vol. 77, No. 396
\newline\texttt{http://orion.math.iastate.edu/dpigozzi/papers/aaldedth.pdf}
\bibitem{Computability}
George S. Boolos, John P. Burgess, Richard C. Jeffrey, (1974).
Computability and Logic. Cambridge University Press.
\bibitem{DecisionProblem}
Egon B\"orger, Erich Gr\"adel, Yuri Gurevich, (2001). The Classical
Decision Problem. Universitext. Springer.
\bibitem{Burris}
Stanley Burris, H.P. Sankappanavar, (1980). A Course in Universal
Alegbra. Springer.
\bibitem{Chang}
C.C. Chang, (1973). Model Theory. Studies in Logic and the
Foundations of Mathematics. Vol 73.
\bibitem{Cirulis}
Cirulis J., (1988). An Algebraization of First-Order Logic With
Terms, in Algebraic Logic (Proc. Conf. Budapest 1988), H. Andr´eka,
J.D. Monk and I. N´emeti, Eds., Colloq. Math. Soc. J. Bolyai,
North-Holland, Amsterdam 1991, pp. 125-146
\bibitem{Cooper}
S. Barry Cooper, (2004). Computability Theory. Chapman \& Hall.
\bibitem{Cori1}
Ren\'e Cori, Daniel Lascar, Donald Pelletier, (2000). Mathematical
Logic, Part 1. Oxford Universty Press.
\bibitem{Cori2}
Ren\'e Cori, Daniel Lascar, Donald Pelletier, (2000). Mathematical
Logic, Part 2. Oxford Universty Press.
\bibitem{Cutland}
Nigel Cutland, (1980). Computability. Cambridge University Press.
\bibitem{Dowek}
Gilles Dowek, (2011). Les D\'emonstrations et les Algorithmes.
Introduction \`a la Logique et \`a la Calculabilit\'e. Editions de
l'Ecole Polytechnique.
\bibitem{Ebbomghaus}
H.D. Ebbinghaus, J. Flum, W. Thomas, (1994). Mathematical Logic.
Springer.
\bibitem{Ferenczi}
Mikl\'os Ferenczi, Mikl\'os Sz\H ots, (2011). Mathematical Logic for
Applications.
\newline\texttt{http://tankonyvtar.ttk.bme.hu/pdf/22.pdf}
\bibitem{Forssell}
Henrik Forssell, (2008). First-Order Logical Duality. PhD
dissertation.
\newline\texttt{http://www.andrew.cmu.edu/user/awodey/students/forssell.pdf}
\bibitem{Forster}
Thomas Forster, (2003). Logic, Induction and Sets. London
Mathematical Society Student Texts.
\bibitem{GabbayDuality}
Murdoch J. Gabbay, (2011). Stone Duality for first-order logic: a
nominal approach to logic and topology. Howard Barringer
Festschrift.
\newline\texttt{http://www.gabbay.org.uk/papers/stodfo.pdf}
\bibitem{GabbaySub}
Murdoch J. Gabbay, (2009). A study of substitution, using nominal
techniques and Fraenkel-Mostowski sets. Theoretical Computer
Science, Volume 410, Issues 12-13, 17 March 2009, Pages 1159-1189.
\newline\texttt{http://www.gabbay.org.uk/papers/stusun.pdf}
\bibitem{GabbayCapture}
Murdoch J. Gabbay and Aad Mathijssen, (2008). Capture-avoiding
Substitution as a Nominal Algebra. Formal Aspects of Computing,
Volume 20, Numbers 4-5, July, 2008, Pages 451-479.
\newline\texttt{http://www.gabbay.org.uk/papers/capasn-jv.pdf}
\bibitem{Gallier}
Jean H. Gallier, (1987). Logic for Computer Science. John Wiley \&
Sons.
\bibitem{Ganesalingam}
Mohan Ganesalingam, (2013). The Language of Mathematics. Springer.
\bibitem{Girard1}
Jean-Yves Girard, (2006). Le Point Aveugle I. Cours de Logique.
Hermann.
\bibitem{Girard2}
Jean-Yves Girard, (2007). Le Point Aveugle II. Cours de Logique.
Hermann.
\bibitem{Halmos}
Paul R. Halmos, (1962). Algebraic Logic. Chelsea Publishing Company
New York.
\bibitem{Hodges}
Wilfrid Hodges, (1993). Model Theory. Cambridge University Press.
\bibitem{Hoyois}
Marc Hoyois, (2012). The Syntax of First Order Logic. Preprint.
\newline\texttt{http://math.northwestern.edu/\~{}hoyois/papers/logic.pdf}
\bibitem{Jech}
Thomas Jech, (2003). Set Theory (3rd ed.). Springer.
\bibitem{Johnstone}
P.T. Johnstone, (1987). Notes on Logic and Set Theory. Cambridge
Mathematical Textbooks. Cambridge University Press.
\bibitem{JohnstoneStone}
P.T. Johnstone, (1986). Stone Spaces. Cambridge Studies in Advanced
Mathematics.
\bibitem{Kanamori}
Akihiro Kanamori, (2009). The Higher Infinite (2nd ed.). Springer.
\bibitem{Kaye}
Richard Kaye, (2007). The Mathematics of Logic. Cambridge University
Press.
\bibitem{Kleene}
Stephen Cole Kleene (2009). Introduction to Metamathematics. Ishi
Press International.
\bibitem{Kohlenbach}
Ulrich Kohlenbach, (2008). Applied Proof Theory: Proof
Interpretations and their Use in Mathematics. Springer.
\bibitem{Krivine}
Jean-Louis Krivine, (2007). Th\'eorie des Ensembles (2nd ed.).
Nouvelle Biblioth\`eque Math\'ematique.
\bibitem{Kunen}
Kenneth Kunen, (2009). The Foundations of Mathematics. Studies in
Logic. Mathematical Logic and Foundations. Vol 19. College
Publications.
\bibitem{KunenSet}
Kenneth Kunen, (2011). Set Theory. Studies in Logic. Mathematical
Logic and Foundations. Vol 34. College Publications.
\bibitem{Marker}
David Marker, (2010). Model Theory: An Introduction. Graduate Texts
in Mathematics. Springer.
\bibitem{Mayberry}
J.P. Mayberry (2012). First Order Logic via the Theory of Functional
Forms.\,\texttt{http://www.bristol.ac.uk/philosophy/department/staff/jpm/logicalt.pdf}
\bibitem{Mendelson}
Elliott Mendelson (2009). Introduction to Mathematical Logic. CRC
Press
\bibitem{Monk}
J. Donald Monk, (1976). Mathematical Logic. Springer.
\bibitem{Metamath}
Norman Megill, Metamath, \texttt{http://www.metamath.org/}
\bibitem{Planet}
PlanetMath, (2012). Deduction theorem holds for first order logic.
\newline\texttt{http://planetmath.org/DeductionTheoremHoldsForFirstOrderLogic.html}
\bibitem{Pohlers}
Wolfram Pohlers, (2009). Proof Theory. Universitext. Springer.
\bibitem{Poizat}
Bruno Poizat, (2000). A Course in Model Theory. Universitext.
Springer.
\bibitem{Rautenberg}
Wolfgang Rautenberg, (2006). A Concise Introduction to Mathematical
Logic. Universitext. Springer.
\bibitem{Rivieccio}
Umberto Rivieccio, (2012). What is Abstract Algebraic Logic?
Preprint.\newline\texttt{http://www.cs.bham.ac.uk/\~{}rivieccu/pub/wha.pdf}
\bibitem{Rudin}
Walter Rudin, (1987). Real and Complex Analysis, 3rd ed.
McGraw-Hill.
\bibitem{Sain}
Ildik\'o Sain, (2000). On the Search for a Finitizable
Algebraization of First Order Logic. Logic Journal of the IGPL
8(4):497-591.
\newline\texttt{http://citeseerx.ist.psu.edu/viewdoc/summary?doi=10.1.1.23.2238}
\bibitem{Schwartz}
Jacob T. Schwartz, Domenico Cantone, Eugenio G. Omodeo, (2011).
Computational Logic and Set Theory. Springer.
\bibitem{Simpson}
Stephen G. Simpson, (2009). Subsystems of Second Order Arithmetic.
Perspective in Logic. Cambridge University Press.
\bibitem{Singh} Arindama Singh, (2012). Mathematical Logic.
Online Course. IIT Madras.
\newline\texttt{http://nptel.iitm.ac.in/courses/111106052/}
\bibitem{SmithGodel}
Peter Smith, (2007). An Introduction to G\"odel's Theorems.
Cambridge University Press.
\bibitem{SmithForm}
Peter Smith, (2003). Formal Logic. Cambridge University Press.
\bibitem{Smullyan}
Raymond M. Smullyan, Melvin Fitting, (1996). Set Theory and the
Continuum Problem. Dover Publications, Inc.
\bibitem{Srivastava}
S.M. Srivastava, (2008). A course on mathematical logic. Springer.
\bibitem{Takeutiproof}
G. Takeuti, (1975). Proof Theory. Dover Publications, Inc.
\bibitem{Takeuti}
G. Takeuti, W.M. Zaring, (1973). Introduction to Axiomatic Set
Theory. Springer.
\bibitem{Tourlakis}
George Tourlakis, (2003). Lectures in Logic and Set Theory. Volume
1. Cambridge University Press.
\bibitem{VanDalen}
Dirk Van Dalen, (2004). Logic and Structure. Universitext. Springer.
\bibitem{Voutsadakis}
George Voutsadakis, (2004). On the Categorical Algebras of First
Order Logic. Scientiae Mathematicae Japonicae Online, Vol. 10, 47–54
\newline\texttt{http://www.voutsadakis.com/RESEARCH/PUBLISHED/cylindric.pdf}
\end{thebibliography}
