\begin{defin}\label{Cat:def:locally:small}
    A category \Cat\ is said to be {\em locally small} \ifand\ the hom-set
    $\Cat(a,b)$ associated with every ordered pair of objects $(a,b)$ is 
    actually a set.
\end{defin}

\begin{prop}\label{Cat:prop:small:locally:small}
    A small category is locally small.
\end{prop}
\begin{proof}
   Let \Cat\ be a small category and $a,b\in\Cat$. We need to show 
    that $\Cat(a,b)$ is a set. Using axiom~(\ref{Cat:ax:collection:set}),
    it is sufficient to find a set $A$ such that every member of
    $\Cat(a,b)$ is an element of $A$. Take $A=\arr\ \Cat$. Then
    from definition~(\ref{Cat:def:category:small}) $A$ is indeed a set,
    and every member of $\Cat(a,b)$ is an element of $A$ by 
    virtue of definition~(\ref{Cat:def:homset}).
\end{proof}

\begin{prop}\label{Cat:prop:locally:small:opposite}
    A category \Cat\ is locally small \ifand\ \Cop is locally small.
\end{prop}
\begin{proof}
    The category \Cat\ being locally small is equivalent to $\Cat(a,b)$
    being a set for all $a,b\in\ob\ \Cat$. Since $\ob\ \Cat=\ob\ \Cop$
    and $\Cop(a,b)=\Cat(b,a)$ from 
    proposition~(\ref{Cat:prop:homset:opposite}), this is in turn equivalent 
    to $\Cop(a,b)$ being a set for all $a,b\in\ob\ \Cop$. Hence, it is
    equivalent to \Cop\ being locally small.
\end{proof}

\begin{prop}\label{Cat:prop:locally:small:product}
    The product $\Cone\times\Ctwo$ of locally small categories
    is locally small.
\end{prop}
\begin{proof}
    Let \Cone\ and \Ctwo\ be two locally small categories. We need to show that
    the canonical product $\Cone\times\Ctwo$ is itself locally small. In other
    words, given $a,b\in\Cone\times\Ctwo$ we need to show that the collection
    $\Cone\times\Ctwo(a,b)$ is actually a set. However from 
    proposition~(\ref{Cat:prop:homset:product}) we have 
    $\Cone\times\Ctwo\,(a,b)=\Cone(a_{1},b_{1})\times\Ctwo(a_{2},b_{2})$
    where $a=(a_{1},a_{2})$ and $b=(b_{1},b_{2})$. So the proposition
    follows from the fact that both $\Cone(a_{1},b_{1})$ and 
    $\Ctwo(a_{2},b_{2})$ are sets, $\Cone$ and $\Ctwo$ being locally 
    small.
\end{proof}
\begin{prop}\label{Cat:prop:locally:small:set}
    The category \Set\ is locally small.
\end{prop}
\begin{proof}
    Given two sets $a,b$, we need to that the hom-set $\Set(a,b)$ is 
    actually a set. Using axiom~(\ref{Cat:ax:collection:set}), it
    is sufficient to show that this collection is a sub-collection
    of another collection which is known to be a set. Now any 
    (untyped) function $f:a\to b$ is a set of ordered pairs $(x,y)$
    where $x\in a$ and $y\in b$ (and with additional properties). 
    In particular, a function $f:a\to b$
    is a subset of the cartesian product $a\times b$, i.e.
    $f\subseteq a\times b$. Equivalently, $f$ is an element of
    the power set ${\cal P}(a\times b)$, i.e. 
    $f\in{\cal P}(a\times b)$. Consequently, we have:
        \begin{eqnarray*}\Set(a,b)
            &=&\{\ (a,b,f)\ |\ \mbox{$f$ is a function $f:a\to b$}\ \}
            \ \leftarrow\ \mbox{prop.~(\ref{Cat:prop:homset:set})}\\
            &\subseteq&\{\ (a,b,f)\ |\ f\in{\cal P}(a\times b)\ \}\\
            &=&\{\ (x,y,f)\ |\ x\in\{a\}\ 
            ,\ y\in\{b\}\ ,\ f\in{\cal P}(a\times b)\ \}\\
            &=&\{a\}\times\{b\}\times{\cal P}(a\times b)
        \end{eqnarray*}
\end{proof}
