\begin{defin}\label{Cat:def:opposite}
    Let $\Cat=(\ob, \arr, \dom,\cod, \id, \circ)$ be a category.
    We call {\em opposite category of \Cat}, the category denoted
    \Cop\ and defined by:
        \[
            \Cop=(\ob, \arr, \cod, \dom, \id, \circ')
        \]
    where the composition operator $\circ'$ is defined by 
    $f \circ' g = g\circ f$, for all $f,g\in\arr$.
\end{defin}

So if $\Cat=(\ob, \arr, \dom,\cod, \id, \circ)$ is a category, the opposite
category \Cop\ is almost identical, except for the composition operator
$\circ'$ which is a flipped version of $\circ$, and for $\dom$ and $\cod$ 
which have been swapped with each other. The collection of objects of
\Cop\ is the same as that of \Cat, giving us the equality $\ob\ \Cop=\ob\ 
\Cat$. Likewise, the collection of arrows of \Cop\ is the same as that of 
\Cat, giving us this other equality $\arr\ \Cop=\arr\ \Cat$. If we denote
$\dom':\arr\to\ob$ and $\cod':\arr\to\ob$ the domain and codomain maps
on \Cop, then $\dom'=\cod$ and $\cod'=\dom$. The identity operator 
$\id:\ob\to\arr$ is the same for both \Cat\ and \Cop, and the composition
arrow $f\circ' g$ in \Cop\ is defined whenever the composition arrow
$g\circ f$ in \Cat\ is defined, and we have $f\circ' g= g \circ f$.

\begin{prop}\label{Cat:prop:opp:is:category}
    Let \Cat\ be a category. Then \Cop\ of 
    definition~(\ref{Cat:def:opposite}) is a category.
\end{prop}
\begin{proof}
    We need to check that the data $\Cop=(\ob, \arr, \cod, \dom, \id, \circ')$ 
    of definition~(\ref{Cat:def:opposite}) forms a category, having assumed
    that the underlying data for \Cat\ does. We have indeed two collections 
    \ob\ and \arr\ with maps between them $\cod:\arr\to\ob$, $\dom:\arr\to\ob$, 
    $\id:\ob\to\arr$ and partial map $\circ':\arr\times\arr\to\arr$. So it
    remains to show that conditions~$(7)-(13)$ of 
    definition~(\ref{Cat:def:category}) are satisfied. For the purpose of
    this proof, we shall denote $\dom'=\cod$ and $\cod'=\dom$.

    $(7)$: We need to check that $f\circ' g$ is defined \ifand\ 
    $\cod'(g)=\dom'(f)$ which is $\dom(g)=\cod(f)$. However by definition, 
    we have set $f\circ' g$ to be defined whenever $g\circ f$ is itself 
    defined, and since \Cat\ is a category, this is in turn equivalent 
    to $\cod(f)=\dom(g)$. Hence, we are done.

    $(8)$: We need to check that $\dom'(f\circ' g)=\dom'(g)$ which can be
    written as $\cod(g\circ f)=\cod(g)$ and which is true since \Cat\ is a
    category.


    $(9)$: We need to check that $\cod'(f\circ' g)=\cod'(f)$ which can be
    written as $\dom(g\circ f)=\dom(f)$ and which is true since \Cat\ is a
    category.

    $(10)$: Given arrows $h,g,f$ with $\cod'(h)=\dom'(g)$ and $\cod'(g)=
    \dom'(f)$, we need to check that $(f\circ' g)\circ' h= f\circ'(g\circ' h)$.
    However, our assumption can be written as $\dom(h)=\cod(g)$ and
    $\dom(g)=\cod(f)$ and having assumed that \Cat\ is a category, by
    property $(10)$ of definition~(\ref{Cat:def:category}) we have:
        \begin{eqnarray*}(f\circ' g)\circ' h 
            &=&h\circ(f\circ' g)\\
            &=&h\circ(g\circ f)\\
            \mbox{\Cat\ is a category\ $\to$\ }
            &=&(h\circ g)\circ f\\
            &=&f\circ' (h\circ g)\\
            &=&f\circ' (g\circ' h)
        \end{eqnarray*}

    $(11)$: We need to check that $\dom'\,(\,\id(a)\,) = a = \cod'
    \,(\,\id(a)\,)$ for all $a\in\Cat$, which follows from $\dom'=\cod$,
    $\cod'=\dom$ and the fact that \Cat\ is a category.

    $(12)$: We need to check that $f\circ'\id(a)=f$ whenever $\dom'(f)=a$,
    that is $\id(a)\circ f=f$ whenever $\cod(f)=a$, which follows from
    \Cat\ being a category.

    $(13)$: We need to check that $\id(a)\circ' f=f$ whenever $\cod'(f)=a$,
    that is $f\circ\id(a)=f$ whenever $\dom(f)=a$, which follows from
    \Cat\ being a category.
\end{proof}

\begin{prop}
    Let \Cat\ be a category. Then the opposite category of \Cop\ is \Cat, i.e.
        \[
            (\Cop)^{op} = \Cat
        \]
\end{prop}
\begin{proof}
    Let $\Cat=(\ob, \arr, \dom,\cod, \id, \circ)$ be a category. From
    definition~(\ref{Cat:def:opposite}), we have
    $\Cop=(\ob, \arr, \cod, \dom, \id, \circ')$ and consequently:
        \[
            (\Cop)^{op}=(\ob, \arr, \dom, \cod, \id, \circ'')
        \]
    In order to show that $(\Cop)^{op}=\Cat$ we simply need to show that
    the partial maps $\circ,\circ'':\arr\times\arr\to\arr$ are equal.
    Given two arrows $f$ and $g$, the composition arrow $g\circ'' f$ is 
    defined \ifand\ $f\circ' g$ is defined, which is itself equivalent to 
    $g\circ f$ being defined. So both $\circ$ and $\circ''$ are defined
    on the same collection of arrow tuples $(g,f)$. Furthemore, whenever
    $g\circ f$ is defined, we have $g\circ'' f=f\circ' g=g\circ f$.
\end{proof}

